---
title: "Exploring the package {climateR}"
author: "Aditya Dahiya"
subtitle: "Exploring the various datasets available publicly and freely to use and analyze with R, {terra} and {sf} with the {climateR} package"
date: "2025-01-30"
bibliography: references.bib
format:
  html:
    code-fold: true
editor_options: 
  chunk_output_type: console
execute: 
  error: false
  message: false
  warning: false
  eval: false
categories:
  - "Maps"
  - "Geocomputation"
  - "{geodata}"
comments:
  giscus: 
    repo: Aditya-Dahiya/visage
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/visage/geocomputation/climateR_package.html"
  description: "Visualizing Information and Spatial Analysis with ggplot2 Extensions"
  twitter: true
  linkedin: true
  email: true
  mastodon: true
  facebook: true
---

## Exploring the {climateR} package

Exploring [{climateR}](https://github.com/mikejohnson51/climateR) (by [Mike Johnson](http://mikejohnson51.github.io/)) with {sf} [@sf-3] and {terra} [@terra]

```{r}
#| label: setup

# Install {cliamteR} package
# remotes::install_github("mikejohnson51/AOI") # suggested!
# remotes::install_github("mikejohnson51/climateR")

# Data Import and Wrangling Tools
library(sf)                   # Handling simple features in R
library(terra)                # Handling rasters in R
library(tidyterra)            # Rasters with ggplot2
library(tidyverse)            # All things tidy

# Final plot tools
library(scales)               # Nice Scales for ggplot2
library(fontawesome)          # Icons display in ggplot2
library(ggtext)               # Markdown text in ggplot2
library(showtext)             # Display fonts in ggplot2
library(colorspace)           # Lighten and Darken colours

# Package to explore
library(climateR)             # Climate Data
library(AOI)                  # Get area of interest

# Making tables in R
library(gt)                   # Beautiful Tables

bts = 12 # Base Text Size
sysfonts::font_add_google("Asap Condensed", "body_font")
showtext::showtext_auto()
theme_set(
  theme_minimal(
    base_size = bts,
    base_family = "body_font"
  ) +
    theme(
      text = element_text(
        colour = "grey20",
        lineheight = 0.3,
        margin = margin(0,0,0,0, "pt")
      )
    )
)

# Some basic caption stuff
# A base Colour
bg_col <- "white"
seecolor::print_color(bg_col)

# Colour for highlighted text
text_hil <- "grey30"
seecolor::print_color(text_hil)

# Colour for the text
text_col <- "grey20"
seecolor::print_color(text_col)


# Caption stuff for the plot
sysfonts::font_add(
  family = "Font Awesome 6 Brands",
  regular = here::here("docs", "Font Awesome 6 Brands-Regular-400.otf")
)
github <- "&#xf09b"
github_username <- "aditya-dahiya"
xtwitter <- "&#xe61b"
xtwitter_username <- "@adityadahiyaias"
social_caption_1 <- glue::glue("<span style='font-family:\"Font Awesome 6 Brands\";'>{github};</span> <span style='color: {text_hil}'>{github_username}  </span>")
social_caption_2 <- glue::glue("<span style='font-family:\"Font Awesome 6 Brands\";'>{xtwitter};</span> <span style='color: {text_hil}'>{xtwitter_username}</span>")
plot_caption <- paste0(
  "**Data**: {geodata} package  ",
  "  |  **Code:** ", 
  social_caption_1, 
  " |  **Graphics:** ", 
  social_caption_2
  )
rm(github, github_username, xtwitter, 
   xtwitter_username, social_caption_1, 
   social_caption_2)
```

## `getTerraClim()`

The `getTerraClim()` function in the `{climateR}` package provides access to TerraClimate data, a high-resolution dataset that offers monthly climate and water balance variables from 1958 onward. These variables include key climate indicators such as precipitation, temperature, soil moisture, wind speed, and evaporation metrics.

```{r}
#| eval: true
#| label: tbl-vars-getTerraClim
#| tbl-cap: "Summary of Climate and Hydrological Variables in getTerraClim()"

terraClim_vars <- tibble::tibble(
  Variable = c("aet", "def", "PDSI", "pet", "ppt", "q", "soil", "srad", 
               "swe", "tmax", "tmin", "vap", "vpd", "ws"),
  Units = c("mm", "mm", "unitless", "mm", "mm", "mm", "mm", "W/m^2", 
            "mm", "degC", "degC", "kPa", "kPa", "m/s"),
  Description = c("Water evaporation amount", 
                  "Potential evaporation minus actual evaporation",
                  "Palmer Drought Severity Index",
                  "Potential evaporation amount",
                  "Precipitation amount",
                  "Runoff amount",
                  "Soil moisture content",
                  "Downwelling shortwave flux in air",
                  "Liquid water content of surface snow",
                  "Maximum air temperature",
                  "Minimum air temperature",
                  "Water vapor partial pressure in air",
                  "Vapor pressure deficit",
                  "Wind speed")
)

terraClim_vars |> 
  gt::gt() |> 
  gtExtras::gt_theme_538()
```

```{r}
#| label: explore-climateR

# Get a nice map of India from Survey of India (simplified to
# save on computing time)
india_vec <- read_sf(
  here::here(
    "data", "india_map",
    "India_Country_Boundary.shp"
  )
) |> 
  # Simplify to save computing time
  st_simplify(dTolerance = 3000) |> 
  # Keep on bigger szed polygons and multiploygons
  filter(!st_is_empty(geometry)) |> 
  arrange(desc(Area)) |> 
  slice(1:2) |> 
  select(-Area) |> 
  st_transform("EPSG:4326")

india_vec |> 
  ggplot() +
  geom_sf()

india_vec2 <- aoi_get(country = "India") |> 
  st_cast("POLYGON")

climate_raw <- getTerraClim(
  india_vec2,
  varname = "ppt",
  startDate = "2021-01-01"
)

ggplot() +
  geom_spatraster(
    data = climate_raw$ppt
  )

object.size(climate_raw)
```

## British Isles: Monthly Rainfall Data

This code generates a faceted map of monthly rainfall in the British Isles from January 2018 to December 2022. It begins by defining the area of interest using [`aoi_get()`](https://rdrr.io/github/adokter/climateR/man/aoi_get.html) and then retrieves precipitation data via [`getTerraClim()`](https://rdrr.io/github/adokter/climateR/man/getTerraClim.html). The region is refined using [`ne_countries()`](https://rdrr.io/cran/rnaturalearth/man/ne_countries.html) and processed with [`st_union()`](https://r-spatial.github.io/sf/reference/st_union.html), [`st_as_sf()`](https://r-spatial.github.io/sf/reference/st_as_sf.html), and [`st_crop()`](https://r-spatial.github.io/sf/reference/st_crop.html) to match the raster data’s extent. The precipitation raster is cropped and masked using [`terra::crop()`](https://rspatial.github.io/terra/reference/crop.html) and [`terra::mask()`](https://rspatial.github.io/terra/reference/mask.html). Monthly labels are extracted with [`str_remove()`](https://stringr.tidyverse.org/reference/str_remove.html) and [`format()`](https://rdrr.io/r/base/format.Date.html). The plot is created using [`ggplot()`](https://ggplot2.tidyverse.org/reference/ggplot.html) with [`geom_spatraster()`](https://rdrr.io/pkg/spatial/pkg/ggraster/man/geom_spatraster.html) for the raster data and [`geom_sf()`](https://ggplot2.tidyverse.org/reference/geom_sf.html) for country borders. Color scaling is handled by [`paletteer::scale_fill_paletteer_c()`](https://rdrr.io/pkg/paletteer/man/scale_fill_paletteer.html), and faceting is applied with [`facet_wrap()`](https://ggplot2.tidyverse.org/reference/facet_wrap.html). The final visualization is styled using [`theme_map()`](https://rdrr.io/pkg/ggthemes/man/theme_map.html), and the plot is saved with [`ggsave()`](https://ggplot2.tidyverse.org/reference/ggsave.html).

```{r}
# Try for a smaller area

uk_vec <- aoi_get(country = c("United Kingdom", "Ireland"))

uk_climate_raw <- getTerraClim(
  uk_vec,
  varname = "ppt",
  startDate = "2018-01-01",
  endDate = "2022-12-31"
)

uk_vec_detail <- rnaturalearth::ne_countries(
  country = c("United Kingdom", "Ireland"),
  scale = "medium",
  returnclass = "sf"
) |> 
  select(geometry) |> 
  st_union() |> 
  st_as_sf() |> 
  st_crop(
    st_bbox(
      c(
        ymin = 49.8,
        ymax = 59.5,
        xmin = -10.3,
        xmax = 2
      ),
      crs = st_crs("EPSG:4326")
    )
  ) |> 
  st_transform(st_crs(uk_climate_raw$ppt))

ggplot(uk_vec_detail) +
  geom_sf()

uk_rast <- uk_climate_raw$ppt |> 
  terra::crop(uk_vec_detail) |> 
  terra::mask(uk_vec_detail)


strip_labels <- uk_rast |> 
  names() |> 
  str_remove("ppt_") |> 
  str_remove("_total") |> 
  as_date() |> 
  format("%B %Y")

names(strip_labels) <- uk_rast |> 
  names()


g <- ggplot() +
  geom_spatraster(
    data = uk_rast[[c(1:60)]]
  ) +
  geom_sf(
    data = uk_vec_detail,
    fill = "transparent",
    linewidth = 0.25,
    colour = "grey20"
  ) +
  paletteer::scale_fill_paletteer_c(
    "grDevices::Blues 3",
    direction = -1,
    na.value = "transparent",
    limits = c(0, 450),
    oob = scales::squish
  ) +
  facet_wrap(
    ~lyr,
    labeller = labeller(lyr = strip_labels),
    nrow = 5,
    ncol = 12
  ) +
  coord_sf(clip = "off") +
  labs(
    title = "British Isles: Rainfall Pattern",
    fill = "Monthly Precipitation (mm)"
  ) +
  ggthemes::theme_map(
    base_size = 40,
    base_family = "body_font"
  ) +
  theme(
    legend.position = "bottom",
    legend.title.position = "top",
    strip.text = element_text(
      hjust = 0,
      margin = margin(4,0,-2,0,"pt"),
      size = 48
    ),
    strip.background = element_blank(),
    plot.title = element_text(
      margin = margin(0,0,10,0, "pt"),
      hjust = 0.5,
      size = 120
    ),
    panel.background = element_rect(
      fill = "transparent",
      colour = "transparent"
    ),
    legend.key.height = unit(10, "pt"),
    legend.key.width = unit(100, "pt"),
    legend.title = element_text(
      margin = margin(0,0,2,0, "mm"),
      size = 80,
      hjust = 0.5
    ),
    legend.text = element_text(
      margin = margin(2,0,0,0, "mm")
    ),
    plot.margin = margin(10,0,10,0, "pt"),
    legend.margin = margin(0,0,0,0, "pt"),
    legend.box.margin = margin(0,0,0,0, "pt"),
    legend.justification = c(0.5, 0)
  )

ggsave(
  filename = here::here(
    "geocomputation", "images",
    "climateR_package_1.png"
  ),
  plot = g,
  height = 4500,
  width = 6000,
  units = "px",
  bg = "white"
)
```

![This faceted visualization illustrates monthly precipitation trends across the British Isles from January 2018 to December 2022. Each panel represents a different month, showcasing variations in rainfall intensity using a gradient color scale. Data is sourced from TerraClimate and spatially refined for accuracy.](images/climateR_package_1.png){#fig-1}
