---
title: "Exploring the package {climateR}"
author: "Aditya Dahiya"
subtitle: "Exploring the various datasets available publicly and freely to use and analyze with R, {terra} and {sf} with the {climateR} package"
date: "2025-01-30"
bibliography: references.bib
format:
  html:
    code-fold: true
editor_options: 
  chunk_output_type: console
execute: 
  error: false
  message: false
  warning: false
  eval: false
categories:
  - "Maps"
  - "Geocomputation"
  - "{geodata}"
comments:
  giscus: 
    repo: Aditya-Dahiya/visage
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/visage/geocomputation/climateR_package.html"
  description: "Visualizing Information and Spatial Analysis with ggplot2 Extensions"
  twitter: true
  linkedin: true
  email: true
  mastodon: true
  facebook: true
---

## Exploring the {climateR} package

Exploring [{climateR}](https://github.com/mikejohnson51/climateR) (by [Mike Johnson](http://mikejohnson51.github.io/)) with {sf} [@sf-3] and {terra} [@terra]

```{r}
#| label: setup

# Install {cliamteR} package
# remotes::install_github("mikejohnson51/AOI") # suggested!
# remotes::install_github("mikejohnson51/climateR")

# Data Import and Wrangling Tools
library(sf)                   # Handling simple features in R
library(terra)                # Handling rasters in R
library(tidyterra)            # Rasters with ggplot2
library(tidyverse)            # All things tidy

# Final plot tools
library(scales)               # Nice Scales for ggplot2
library(fontawesome)          # Icons display in ggplot2
library(ggtext)               # Markdown text in ggplot2
library(showtext)             # Display fonts in ggplot2
library(colorspace)           # Lighten and Darken colours

# Package to explore
library(climateR)             # Climate Data
library(AOI)                  # Get area of interest

# Making tables in R
library(gt)                   # Beautiful Tables

bts = 12 # Base Text Size
sysfonts::font_add_google("Asap Condensed", "body_font")
showtext::showtext_auto()
theme_set(
  theme_minimal(
    base_size = bts,
    base_family = "body_font"
  ) +
    theme(
      text = element_text(
        colour = "grey20",
        lineheight = 0.3,
        margin = margin(0,0,0,0, "pt")
      )
    )
)

# Some basic caption stuff
# A base Colour
bg_col <- "white"
seecolor::print_color(bg_col)

# Colour for highlighted text
text_hil <- "grey30"
seecolor::print_color(text_hil)

# Colour for the text
text_col <- "grey20"
seecolor::print_color(text_col)


# Caption stuff for the plot
sysfonts::font_add(
  family = "Font Awesome 6 Brands",
  regular = here::here("docs", "Font Awesome 6 Brands-Regular-400.otf")
)
github <- "&#xf09b"
github_username <- "aditya-dahiya"
xtwitter <- "&#xe61b"
xtwitter_username <- "@adityadahiyaias"
social_caption_1 <- glue::glue("<span style='font-family:\"Font Awesome 6 Brands\";'>{github};</span> <span style='color: {text_hil}'>{github_username}  </span>")
social_caption_2 <- glue::glue("<span style='font-family:\"Font Awesome 6 Brands\";'>{xtwitter};</span> <span style='color: {text_hil}'>{xtwitter_username}</span>")
plot_caption <- paste0(
  "**Data**: {climateR} package  ",
  "  |  **Code:** ", 
  social_caption_1, 
  " |  **Graphics:** ", 
  social_caption_2
  )
rm(github, github_username, xtwitter, 
   xtwitter_username, social_caption_1, 
   social_caption_2)
```

## `getTerraClim()`

The `getTerraClim()` function in the `{climateR}` package provides access to TerraClimate data, a high-resolution dataset that offers monthly climate and water balance variables from 1958 onward. These variables include key climate indicators such as precipitation, temperature, soil moisture, wind speed, and evaporation metrics.

```{r}
#| eval: true
#| label: tbl-vars-getTerraClim
#| tbl-cap: "Summary of Climate and Hydrological Variables in getTerraClim()"

terraClim_vars <- tibble::tibble(
  Variable = c("aet", "def", "PDSI", "pet", "ppt", "q", "soil", "srad", 
               "swe", "tmax", "tmin", "vap", "vpd", "ws"),
  Units = c("mm", "mm", "unitless", "mm", "mm", "mm", "mm", "W/m^2", 
            "mm", "degC", "degC", "kPa", "kPa", "m/s"),
  Description = c("Water evaporation amount", 
                  "Potential evaporation minus actual evaporation",
                  "Palmer Drought Severity Index",
                  "Potential evaporation amount",
                  "Precipitation amount",
                  "Runoff amount",
                  "Soil moisture content",
                  "Downwelling shortwave flux in air",
                  "Liquid water content of surface snow",
                  "Maximum air temperature",
                  "Minimum air temperature",
                  "Water vapor partial pressure in air",
                  "Vapor pressure deficit",
                  "Wind speed")
)

terraClim_vars |> 
  gt::gt() |> 
  gtExtras::gt_theme_538()
```

## British Isles: Monthly Rainfall Data

This code generates a faceted map of monthly rainfall in the British Isles from January 2018 to December 2022. It begins by defining the area of interest using [`aoi_get()`](https://rdrr.io/github/adokter/climateR/man/aoi_get.html) and then retrieves precipitation data via [`getTerraClim()`](https://rdrr.io/github/adokter/climateR/man/getTerraClim.html). The region is refined using [`ne_countries()`](https://rdrr.io/cran/rnaturalearth/man/ne_countries.html) and processed with [`st_union()`](https://r-spatial.github.io/sf/reference/st_union.html), [`st_as_sf()`](https://r-spatial.github.io/sf/reference/st_as_sf.html), and [`st_crop()`](https://r-spatial.github.io/sf/reference/st_crop.html) to match the raster data’s extent. The precipitation raster is cropped and masked using [`terra::crop()`](https://rspatial.github.io/terra/reference/crop.html) and [`terra::mask()`](https://rspatial.github.io/terra/reference/mask.html). Monthly labels are extracted with [`str_remove()`](https://stringr.tidyverse.org/reference/str_remove.html) and [`format()`](https://rdrr.io/r/base/format.Date.html). The plot is created using [`ggplot()`](https://ggplot2.tidyverse.org/reference/ggplot.html) with [`geom_spatraster()`](https://rdrr.io/pkg/spatial/pkg/ggraster/man/geom_spatraster.html) for the raster data and [`geom_sf()`](https://ggplot2.tidyverse.org/reference/geom_sf.html) for country borders. Color scaling is handled by [`paletteer::scale_fill_paletteer_c()`](https://rdrr.io/pkg/paletteer/man/scale_fill_paletteer.html), and faceting is applied with [`facet_wrap()`](https://ggplot2.tidyverse.org/reference/facet_wrap.html). The final visualization is styled using [`theme_map()`](https://rdrr.io/pkg/ggthemes/man/theme_map.html), and the plot is saved with [`ggsave()`](https://ggplot2.tidyverse.org/reference/ggsave.html).

```{r}
# Try for a smaller area

uk_vec <- aoi_get(country = c("United Kingdom", "Ireland"))

uk_climate_raw <- getTerraClim(
  uk_vec,
  varname = "ppt",
  startDate = "2018-01-01",
  endDate = "2022-12-31"
)

uk_vec_detail <- rnaturalearth::ne_countries(
  country = c("United Kingdom", "Ireland"),
  scale = "medium",
  returnclass = "sf"
) |> 
  select(geometry) |> 
  st_union() |> 
  st_as_sf() |> 
  st_crop(
    st_bbox(
      c(
        ymin = 49.8,
        ymax = 59.5,
        xmin = -10.3,
        xmax = 2
      ),
      crs = st_crs("EPSG:4326")
    )
  ) |> 
  st_transform(st_crs(uk_climate_raw$ppt))

ggplot(uk_vec_detail) +
  geom_sf()

uk_rast <- uk_climate_raw$ppt |> 
  terra::crop(uk_vec_detail) |> 
  terra::mask(uk_vec_detail)


strip_labels <- uk_rast |> 
  names() |> 
  str_remove("ppt_") |> 
  str_remove("_total") |> 
  as_date() |> 
  format("%B\n%Y")

names(strip_labels) <- uk_rast |> 
  names()


g <- ggplot() +
  geom_spatraster(
    data = uk_rast[[c(1:60)]]
  ) +
  geom_sf(
    data = uk_vec_detail,
    fill = "transparent",
    linewidth = 0.25,
    colour = "grey20"
  ) +
  paletteer::scale_fill_paletteer_c(
    "grDevices::Blues 3",
    direction = -1,
    na.value = "transparent",
    limits = c(0, 450),
    oob = scales::squish
  ) +
  facet_wrap(
    ~lyr,
    labeller = labeller(lyr = strip_labels),
    nrow = 5,
    ncol = 12
  ) +
  coord_sf(clip = "off") +
  labs(
    title = "British Isles: Rainfall Pattern",
    fill = "Monthly Precipitation (mm)"
  ) +
  ggthemes::theme_map(
    base_size = 40,
    base_family = "body_font"
  ) +
  theme(
    legend.position = "bottom",
    legend.title.position = "top",
    strip.text = element_text(
      hjust = 0,
      margin = margin(6,0,-10,0,"pt"),
      size = 48,
      lineheight = 0.3
    ),
    strip.background = element_blank(),
    plot.title = element_text(
      margin = margin(0,0,10,0, "pt"),
      hjust = 0.5,
      size = 120
    ),
    panel.background = element_rect(
      fill = "transparent",
      colour = "transparent"
    ),
    legend.key.height = unit(10, "pt"),
    legend.key.width = unit(100, "pt"),
    legend.title = element_text(
      margin = margin(0,0,2,0, "mm"),
      size = 80,
      hjust = 0.5
    ),
    legend.text = element_text(
      margin = margin(2,0,0,0, "mm")
    ),
    plot.margin = margin(10,0,10,0, "pt"),
    legend.margin = margin(0,0,0,0, "pt"),
    legend.box.margin = margin(0,0,0,0, "pt"),
    legend.justification = c(0.5, 0)
  )

ggsave(
  filename = here::here(
    "geocomputation", "images",
    "climateR_package_1.png"
  ),
  plot = g,
  height = 4500,
  width = 6000,
  units = "px",
  bg = "white"
)
```

![This faceted visualization illustrates monthly precipitation trends across the British Isles from January 2018 to December 2022. Each panel represents a different month, showcasing variations in rainfall intensity using a gradient color scale. Data is sourced from TerraClimate and spatially refined for accuracy.](images/climateR_package_1.png){#fig-1}

## Precipitation data graphic for India.

```{r}

# Get a nice map of India from Survey of India (simplified to
# save on computing time)
india_vec <- read_sf(
  here::here(
    "data", "india_map",
    "India_Country_Boundary.shp"
  )
) |> 
  # Simplify to save computing time
  st_simplify(dTolerance = 3000) |> 
  # Keep on bigger szed polygons and multiploygons
  filter(!st_is_empty(geometry)) |> 
  arrange(desc(Area)) |> 
  slice(1:2) |> 
  select(-Area) |> 
  st_transform("EPSG:4326")

# Quick check
india_vec |> 
  ggplot() +
  geom_sf()
#--------------------------------------------------------------
# Minor manual correction: Increase bounding box of India
# as the top of north is slightly getting cropped
# Get the current bounding box of India
india_bbox <- st_bbox(india_vec)

# Increase the northern boundary (ymax) by 2 degrees
india_bbox["ymax"] <- india_bbox["ymax"] + 2

# Convert the updated bounding box to an sf object
india_bbox_sf <- st_as_sfc(india_bbox, crs = st_crs(india_vec))

# Crop india_vec to match the expanded bounding box
india_vec <- st_crop(india_vec, india_bbox_sf)

# Remove temporary objects
rm(india_bbox, india_bbox_sf)
#-------------------------------------------------------------

# Download the TerraClimate dataset on PDSI from 
# https://www.climatologylab.org/terraclimate.html
india_rast_raw <- getTerraClim(
  india_vec,
  varname = "ppt",
  startDate = "2018-01-01",
  endDate = "2023-12-01"
)

# Match CRS of both: prefer to change CRS of vector
india_vec2 <- india_vec |> 
  st_transform(crs(india_rast_raw$ppt))

# Crop and Maskt he Raster to show only the data within 
# India's Administrative Borders
india_rast <- india_rast_raw$ppt |> 
  aggregate(fact = 4) |> 
  terra::crop(india_vec2) |> 
  terra::mask(india_vec2)

# Clean Names for the Panel Strip Text: Month and Year
strip_labels <- india_rast |> 
  names() |> 
  str_remove("ppt_") |> 
  str_remove("_total") |> 
  as_date() |> 
  format("%b\n%Y")

names(strip_labels) <- india_rast |> 
  names()

length(strip_labels)

# Temporarily save india_rast to save download time
# Check its range of values
writeRaster(
  india_rast,
  filename = "india_ppt_rast.tif",
  overwrite = TRUE
)

india_rast |>
  values() |>
  range(na.rm = T)

g2 <- ggplot() +
  geom_spatraster(
    data = india_rast
  ) +
  geom_sf(
    data = india_vec2,
    fill = "transparent",
    linewidth = 0.25,
    colour = "grey20"
  ) +
  paletteer::scale_fill_paletteer_c(
    "grDevices::Blues 3",
    direction = -1,
    na.value = "transparent",
    limits = c(0, 700),
    breaks = seq(0, 600, 200),
    oob = scales::squish
  ) +
  facet_wrap(
    ~lyr,
    labeller = labeller(lyr = strip_labels),
    ncol = 12
  ) +
  coord_sf(clip = "off") +
  labs(
    title = "India: Rainfall Patterns (2018 - 2023)",
    subtitle = str_wrap("Monthly rainfall data for India, from TerraClimate dataset, fetched as a Raster using {climateR}", 130),
    fill = "Precipitation (mm)",
    caption = plot_caption
  ) +
  ggthemes::theme_map(
    base_size = 40,
    base_family = "body_font"
  ) +
  theme(
    legend.position = "bottom",
    legend.title.position = "top",
    strip.text = element_text(
      hjust = 0.7,
      margin = margin(3,0,-20,0,"pt"),
      size = 48,
      lineheight = 0.3
    ),
    strip.background = element_blank(),
    plot.title = element_text(
      margin = margin(0,0,10,0, "pt"),
      hjust = 0.5,
      size = 160
    ),
    plot.subtitle = element_text(
      margin = margin(0,0,10,0, "pt"),
      hjust = 0.5,
      size = 70,
      lineheight = 0.3
    ),
    plot.caption = element_textbox(
      hjust = 0.5,
      margin = margin(30,0,0,0, "pt"),
      size = 60
    ),
    panel.background = element_rect(
      fill = "transparent",
      colour = "transparent"
    ),
    legend.key.height = unit(10, "pt"),
    legend.key.width = unit(100, "pt"),
    legend.title = element_text(
      margin = margin(0,0,2,0, "mm"),
      size = 80,
      hjust = 0.5
    ),
    legend.text = element_text(
      margin = margin(2,0,0,0, "mm"),
      size = 60
    ),
    plot.margin = margin(10,5,5,5, "pt"),
    legend.margin =  margin(-20,0,0,0, "pt"),
    legend.box.margin = margin(-20,0,0,0, "pt"),
    legend.justification = c(0.5, 0)
  )

ggsave(
  filename = here::here(
    "geocomputation", "images",
    "climateR_package_3.png"
  ),
  plot = g2,
  height = 3700,
  width = 4000,
  units = "px",
  bg = "white",
  limitsize = FALSE
)
```

![This faceted visualization illustrates monthly precipitation trends across the India from January 2018 to December 2023. Each panel represents a different month, showcasing variations in rainfall intensity using a gradient color scale. Data is sourced from TerraClimate and spatially refined for accuracy.](images/climateR_package_3.png)

## India: Drought Severity Trends (2010 - 2023)

The Palmer Drought Severity Index ([PDSI](https://climatedataguide.ucar.edu/climate-data/palmer-drought-severity-index-pdsi)) is a widely used metric for measuring long-term drought conditions, based on temperature, precipitation, and soil moisture balance. It provides a standardized scale where negative values indicate drought (with -4 or lower signifying extreme drought) and positive values represent wet conditions.

The analysis shown in @fig-2 uses the [TerraClimate](https://www.climatologylab.org/terraclimate.html) dataset from the [Climatology Lab](https://www.climatologylab.org/), a high-resolution global dataset that offers monthly climate and hydrological variables at a \~4 km resolution. The dataset is accessed via the [`climateR`](https://github.com/mikejohnson51/climateR) package, which facilitates retrieval of climate data from TerraClimate. The geographic boundaries for India are obtained using [Survey of India](https://onlinemaps.surveyofindia.gov.in/Digital_Product_Show.aspx), and spatial data operations are conducted with [`sf`](https://r-spatial.github.io/sf/) and [`terra`](https://rspatial.github.io/terra/). The final visualization employs [`ggplot2`](https://ggplot2.tidyverse.org/) adn [`tidyterra`](https://dieghernan.github.io/tidyterra/) to create a faceted graph displaying PDSI trends from 2010 to 2024, offering insights into regional drought patterns over time.

```{r}

# Get a nice map of India from Survey of India (simplified to
# save on computing time)
india_vec <- read_sf(
  here::here(
    "data", "india_map",
    "India_Country_Boundary.shp"
  )
) |> 
  # Simplify to save computing time
  st_simplify(dTolerance = 3000) |> 
  # Keep on bigger szed polygons and multiploygons
  filter(!st_is_empty(geometry)) |> 
  arrange(desc(Area)) |> 
  slice(1:2) |> 
  select(-Area) |> 
  st_transform("EPSG:4326")

# Quick check
india_vec |> 
  ggplot() +
  geom_sf()

# Minor manual correction: Increase bounding box of India
# as the top of north is slightly getting cropped
# Get the current bounding box of India
india_bbox <- st_bbox(india_vec)

# Increase the northern boundary (ymax) by 2 degrees
india_bbox["ymax"] <- india_bbox["ymax"] + 2

# Convert the updated bounding box to an sf object
india_bbox_sf <- st_as_sfc(india_bbox, crs = st_crs(india_vec))

# Crop india_vec to match the expanded bounding box
india_vec <- st_crop(india_vec, india_bbox_sf)

# Remove temporary objects
rm(india_bbox, india_bbox_sf)

# Download the TerraClimate dataset on PDSI from 
# https://www.climatologylab.org/terraclimate.html
india_rast_raw <- getTerraClim(
  india_vec,
  varname = "PDSI",
  startDate = "2010-01-01",
  endDate = "2023-12-01"
)

# Match CRS of both: prefer to change CRS of vector
india_vec2 <- india_vec |> 
  st_transform(crs(india_rast_raw$PDSI))

# Crop and Maskt he Raster to show only the data within 
# India's Administrative Borders
india_rast <- india_rast_raw$PDSI |> 
  aggregate(fact = 4) |> 
  terra::crop(india_vec2) |> 
  terra::mask(india_vec2)

# Clean Names for the Panel Strip Text: Month and Year
strip_labels <- india_rast |> 
  names() |> 
  str_remove("PDSI_") |> 
  str_remove("_total") |> 
  as_date() |> 
  format("%b\n%Y")

names(strip_labels) <- india_rast |> 
  names()

length(strip_labels)

# Temporarily save india_rast to save download time
# Check its range of values
# writeRaster(
#   india_rast,
#   filename = "india_rast.tif"
# )
# 
# india_rast |> 
#   values() |> 
#   range(na.rm = T)

g2 <- ggplot() +
  geom_spatraster(
    data = india_rast
  ) +
  geom_sf(
    data = india_vec2,
    fill = "transparent",
    linewidth = 0.25,
    colour = "grey20"
  ) +
  paletteer::scale_fill_paletteer_c(
    "ggthemes::Orange-Blue-White Diverging",
    direction = 1,
    na.value = "transparent",
    limits = c(-10, 10),
    oob = scales::squish
  ) +
  facet_wrap(
    ~lyr,
    labeller = labeller(lyr = strip_labels),
    ncol = 12
  ) +
  coord_sf(clip = "off") +
  labs(
    title = "India: Drought Severity (2010 - 2023)",
    subtitle = str_wrap("Palmer Drought Severity Index (PDSI) measures long-term drought and wetness conditions based on temperature, precipitation, and soil moisture balance. Negative values indicate drought severity, while positive values signify wetter-than-normal conditions.", 130),
    fill = "Palmer Drought Severity Index (PDSI)",
    caption = plot_caption
  ) +
  ggthemes::theme_map(
    base_size = 40,
    base_family = "body_font"
  ) +
  theme(
    legend.position = "bottom",
    legend.title.position = "top",
    strip.text = element_text(
      hjust = 0.7,
      margin = margin(3,0,-20,0,"pt"),
      size = 48,
      lineheight = 0.3
    ),
    strip.background = element_blank(),
    plot.title = element_text(
      margin = margin(0,0,10,0, "pt"),
      hjust = 0.5,
      size = 210
    ),
    plot.subtitle = element_text(
      margin = margin(0,0,10,0, "pt"),
      hjust = 0.5,
      size = 90,
      lineheight = 0.3
    ),
    plot.caption = element_textbox(
      hjust = 0.5,
      margin = margin(10,0,0,0, "pt"),
      size = 80
    ),
    panel.background = element_rect(
      fill = "transparent",
      colour = "transparent"
    ),
    legend.key.height = unit(10, "pt"),
    legend.key.width = unit(100, "pt"),
    legend.title = element_text(
      margin = margin(0,0,2,0, "mm"),
      size = 80,
      hjust = 0.5
    ),
    legend.text = element_text(
      margin = margin(2,0,0,0, "mm"),
      size = 60
    ),
    plot.margin = margin(10,5,10,5, "pt"),
    legend.margin =  margin(0,0,0,0, "pt"),
    legend.box.margin = margin(0,0,0,0, "pt"),
    legend.justification = c(0.5, 0)
  )

g1 <- tibble(year = 2010:2023) |> 
  ggplot() +
  geom_text(
    mapping = aes(
      x = 0, y = 0,
      label = year
    ),
    family = "body_font",
    size = 80,
    colour = "grey10"
  ) +
  coord_cartesian(clip = "off") +
  facet_wrap(~year, ncol = 1) +
  theme_void() +
  theme(
    strip.text = element_blank()
  )

library(patchwork)
g <- wrap_plots(g1, g2) +
  plot_layout(widths = c(1, 12))

ggsave(
  filename = here::here(
    "geocomputation", "images",
    "climateR_package_2.png"
  ),
  plot = g2,
  height = 2500 * 4.1,
  width = 6000,
  units = "px",
  bg = "white",
  limitsize = FALSE
)
```

![This faceted visualization displays the monthly PDSI values across India from 2010 to 2024. Each horizontal row represents an year, while each column represents a month, highlighting spatial variations in drought and wetness conditions. Negative values (shades of red) indicate drought severity, while positive values (shades of blue) reflect wetter-than-normal conditions. Data sourced from TerraClimate via the Climatology Lab.](images/climateR_package_2.png){#fig-2}
