---
title: "Using data.gov.in data on PIN Code boundaries in India"
author: "Aditya Dahiya"
subtitle: "Combining {sf}, {terra}, {tidyterra} and {osmdata} to make beautiful postal code maps"
date: "2025-05-10"
bibliography: references.bib
format:
  html:
    code-fold: true
editor_options: 
  chunk_output_type: console
execute: 
  error: false
  message: false
  warning: false
  eval: false
categories:
  - "Maps"
  - "Open Street Maps"
  - "data.gov.in"
  - "{terra}"
  - "tidyterra"
  - "India"
  - "Haryana"
  - "Chandigarh"
comments:
  giscus: 
    repo: Aditya-Dahiya/visage
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/visage/geocomputation/postal_code_maps.html"
  description: "Visualizing Information and Spatial Analysis with ggplot2 Extensions"
  twitter: true
  linkedin: true
  email: true
  mastodon: true
  facebook: true
---

### Load packages

```{r}
#| label: setup
#| eval: true

pacman::p_load(
  sf,
  terra, 
  tidyterra,
  tidyverse,
  showtext,
  ggtext,
  fontawesome,
  geodata,
  ggmap
)
```

### Getting data

The **All India Pincode Boundary GeoJSON dataset**, available on <https://www.data.gov.in/catalog/all-india-pincode-boundary-geo-json>, provides comprehensive geospatial information delineating postal code boundaries across India. This open data resource, published by the [Government of India](https://www.india.gov.in/), includes GeoJSON files that map pincode regions with associated attributes such as pincode numbers, district names, and state names. The dataset is designed to facilitate applications in urban planning, logistics, demographic analysis, and geospatial research. Maintained under the [Open Government Data (OGD) Platform India](https://www.data.gov.in/), it ensures accessibility and transparency, allowing developers, researchers, and policymakers to leverage accurate and standardized pincode-level geospatial data for various analytical and operational purposes.

```{r}
#| eval: true
#| label: get-data

india_pin_code <- read_sf(
  here::here(
    "data",
    "india_map",
    "All_India_pincode_Boundary-19312.geojson"
  )
) |> 
  janitor::clean_names() |> 
  st_transform("EPSG:4326") |> 
  mutate(pincode = parse_number(pincode))

haryana_map <- read_sf(
  here::here(
    "data",
    "haryana_map",
    "HARYANA_DISTRICT_BDY.shp"
  )
) |> 
  janitor::clean_names() |> 
  mutate(
    district = str_replace_all(district, ">", "A"),
    district = str_replace_all(district, "\\|", "I"),
    district = str_to_title(district)
  ) |> 
  st_transform("EPSG:4326")

haryana_outline <- read_sf(
  here::here(
    "data",
    "haryana_map",
    "HARYANA_STATE_BDY.shp"
  )
) |> 
  janitor::clean_names() |> 
  st_simplify(dTolerance = 1000) |> 
  st_transform("EPSG:4326")

chandigarh_postal <- india_pin_code |> 
  filter(pincode >= 160001 & pincode < 160100) |> 
  mutate(
    label_name = str_remove_all(office_name, "SO"),
    label_name = str_remove_all(label_name, "Chandigarh"),
    label_name = paste(
      "<b>", label_name, "</b><br><span style='font-size:12pt'>", 
      pincode,
      "</span>"
    )
  ) |> 
  relocate(label_name)


# register_stadiamap("YOUR KEY HERE")
get_map_bbox <- chandigarh_postal |> 
  st_bbox()
names(get_map_bbox) <- c("left", "bottom", "right", "top")

base_map <- get_stadiamap(
  bbox = get_map_bbox,
  zoom = 13,
  maptype = "stamen_toner_background"
) |> 
  terra::rast()

# ggplot() +
#   geom_spatraster_rgb(data = base_map)

# Get post offices in Chandigarh
pacman::p_load(osmdata)

chandigarh_postoffices <- opq(bbox = st_bbox(chandigarh_postal)) |> 
  add_osm_feature(
    key = "amenity",
    value = c("post_office", "post_box", "post_depot")
  ) |> 
  osmdata_sf()


```

### Data Wrangling and Testing Code

```{r}
  
haryana_outline |> 
  ggplot() +
  geom_sf()

haryana_map |> 
  st_drop_geometry() |> 
  select(district) |> 
  print(n = Inf)

india_pin_code |> 
  st_drop_geometry() |> 
  count(region, sort = T)
  
st_bbox(haryana_outline)

ggplot() +
  geom_sf() +
  geom_sf_text(
    aes(label = str_wrap(office_name, 5)),
    size = 2,
    lineheight = 0.7
  )
```

### Visualization Parameters

```{r}
# Visualization Parameters

bts = 12 # Base Text Size
sysfonts::font_add_google("Saira Condensed", "body_font")
sysfonts::font_add_google("Saira", "title_font")
sysfonts::font_add_google("Saira Extra Condensed", "caption_font")
showtext::showtext_auto()
# A base Colour
bg_col <- "white"
seecolor::print_color(bg_col)

# Colour for highlighted text
text_hil <- "grey30"
seecolor::print_color(text_hil)

# Colour for the text
text_col <- "grey20"
seecolor::print_color(text_col)

theme_set(
  theme_minimal(
    base_size = bts,
    base_family = "body_font"
  ) +
    theme(
      text = element_text(
        colour = "grey30",
        lineheight = 0.3,
        margin = margin(0,0,0,0, "pt")
      ),
      plot.title = element_text(
        hjust = 0.5
      ),
      plot.subtitle = element_text(
        hjust = 0.5
      )
    )
)

# Caption stuff for the plot
sysfonts::font_add(
  family = "Font Awesome 6 Brands",
  regular = here::here("docs", "Font Awesome 6 Brands-Regular-400.otf")
)
github <- "&#xf09b"
github_username <- "aditya-dahiya"
xtwitter <- "&#xe61b"
xtwitter_username <- "@adityadahiyaias"
social_caption_1 <- glue::glue("<span style='font-family:\"Font Awesome 6 Brands\";'>{github};</span> <span style='color: {text_hil}'>{github_username}  </span>")
social_caption_2 <- glue::glue("<span style='font-family:\"Font Awesome 6 Brands\";'>{xtwitter};</span> <span style='color: {text_hil}'>{xtwitter_username}</span>")
plot_caption <- paste0(
  "**Data**:  data.gov.in & Stadia Maps",
  "  |  **Code:** ", 
  social_caption_1, 
  " |  **Graphics:** ", 
  social_caption_2
  )
rm(github, github_username, xtwitter, 
   xtwitter_username, social_caption_1, 
   social_caption_2)

```

### Plotting a static map

This map of postal zones and post office boundaries in Chandigarh, India, was crafted using R and a suite of powerful packages for spatial data manipulation and visualization. The process started by loading key packages with [`pacman::p_load()`](https://cran.r-project.org/web/packages/pacman/), including [**`sf`**](https://r-spatial.github.io/sf/) for handling vector spatial data, [**`terra`**](https://rspatial.github.io/terra/) for raster data support, [**`tidyterra`**](https://dieghernan.github.io/tidyterra/) for integrating raster data with ggplot2, [**`tidyverse`**](https://www.tidyverse.org/) for data wrangling, [**`ggmap`**](https://github.com/dkahle/ggmap) for fetching base maps, [**`showtext`**](https://cran.r-project.org/web/packages/showtext/) for custom fonts, and [**`ggtext`**](https://wilkelab.org/ggtext/) for rich text labeling. I imported the All India Pincode Boundary GeoJSON data using [**`read_sf()`**](https://r-spatial.github.io/sf/reference/st_read.html) from the `sf` package, transformed it to the EPSG:4326 coordinate system with [**`st_transform()`**](https://r-spatial.github.io/sf/reference/st_transform.html), and filtered it to include only Chandigarh’s postal codes (160001 to 160099). A base map was retrieved from [Stadia Maps](https://stadiamaps.com/) via [**`get_stadiamap()`**](https://www.rdocumentation.org/packages/ggmap/versions/3.0.0/topics/get_stadiamap), using the bounding box of the Chandigarh postal data. The map was then built with [**`ggplot2`**](https://ggplot2.tidyverse.org/), layering the base map raster with [**`geom_spatraster_rgb()`**](https://dieghernan.github.io/tidyterra/reference/geom_spatraster_rgb.html), postal boundaries with [**`geom_sf()`**](https://ggplot2.tidyverse.org/reference/ggsf.html), and post office labels with [**`geom_richtext()`**](https://wilkelab.org/ggtext/reference/geom_richtext.html). Custom fonts were applied using `showtext`, and the plot was polished with a minimal theme and saved as a high-resolution PNG using [**`ggsave()`**](https://ggplot2.tidyverse.org/reference/ggsave.html).

```{r}

size_var = 2000
bts = size_var / 30

# nrow(chandigarh_postal)
# cols4all::c4a_gui()
# cols4all::c4a("poly.dark24")


g <- ggplot() +
  geom_spatraster_rgb(
    data = base_map,
    maxcell = Inf,
    alpha = 0.75
  ) +
  geom_sf(
    data = chandigarh_postal,
    mapping = aes(fill = office_name),
    colour = "transparent",
    linewidth = 2,
    alpha = 0.4
  ) +
  ggtext::geom_richtext(
    data = chandigarh_postal,
    mapping = aes(
      label = label_name,
      geometry = geometry
    ),
    fill = alpha("white", 0.5),
    colour = text_col,
    lineheight = 0,
    size = bts / 8,
    hjust = 0.5, 
    vjust = 0.5,
    family = "body_font",
    stat = "sf_coordinates",
    label.size = NA,
    label.padding = unit(0.05, "lines")
  ) +
  # geom_sf(
  #   data = chandigarh_postoffices$osm_points,
  #   size = 5
  # ) +
  cols4all::scale_fill_discrete_c4a_cat("poly.dark24") +
  labs(
    x = NULL, y = NULL,
    title = "Postal Map of Chandigarh (India)",
    subtitle = "Crafted from India's Open Government Data at *data.gov.in*, this map of Chandigarh's postal zones leverages **{sf}**, **{terra}**, and **{tidyterra}**." |> str_wrap(80) |> str_replace_all("\n", "<br>"),
    caption = plot_caption
  ) +
  coord_sf(
    crs = "EPSG:3857",
    expand = FALSE
  ) +
  theme_minimal(
    base_size = bts,
    base_family = "body_font",
    base_line_size = bts / 140,
    base_rect_size = bts / 140
  ) +
  theme(
    
    # Plot / Overall
    plot.margin = margin(5,5,5,5, "pt"),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    text = element_text(
      colour = text_hil,
      lineheight = 0.3,
      margin = margin(0,0,0,0, "pt")
    ),
    
    # Plot Title, Subtitle and Caption
    plot.title = element_text(
      margin = margin(20,0,0,0, "pt"),
      hjust = 0.5,
      size = bts * 1.8
    ),
    plot.subtitle = element_textbox(
      margin = margin(0,0,10,0, "pt"),
      hjust = 0.5,
      halign = 0.5,
      size = bts / 1.4
    ),
    plot.caption = element_textbox(
      margin = margin(5,0,0,0, "pt"),
      halign = 0.5,
      hjust = 0.5,
      size = 0.4 * bts,
      family = "caption_font"
    ),
    axis.text.x = element_text(
      size = bts / 2,
      margin = margin(-0.5,0,0,0, "pt")
    ),
    axis.text.y = element_text(
      size = bts / 2,
      margin = margin(0,-0.5,0,0, "pt")
    ),
    axis.ticks.length = unit(0, "pt"),
    axis.ticks = element_blank(),
    
    # Legend
    legend.position = "none",
    
    # Panel
    panel.grid = element_line(
      linewidth = 0.3,
      colour = "grey30"
    )
  )


ggsave(
  plot = g,
  filename = here::here("geocomputation", "images",
                        "postal_code_maps_1.png"),
  width = size_var,
  height = (5/4) * size_var,
  units = "px",
  bg = bg_col
)

```

![This map visualizes the postal zones and post office boundaries of Chandigarh, India, using open geospatial data from data.gov.in. Each zone is color-coded by post office, with labels displaying office names and pincodes, crafted with {sf}, {terra}, and {ggtext} for precision. A Stadia Maps base layer, fetched via {ggmap}, enhances context, while {ggplot2} and {tidyterra} ensure a polished output.](images/postal_code_maps_1.png){#fig-1}
