---
title: "Chapter 9: Making maps with R"
subtitle: "Key Learnings from, and Solutions to the exercises in Chapter 8 of the book Geocomputation with R by Robin Lovelace, Jakub Nowosad and Jannes Muenchow."
date: "2025-03-02"
author: "Aditya Dahiya"
bibliography: references.bib
format:
  html:
    code-fold: true
editor_options: 
  chunk_output_type: console
execute: 
  error: false
  message: false
  warning: false
  eval: false
  cache: false
categories:
  - "Geocomputation with R"
  - "Textbook Solutions"
comments:
  giscus: 
    repo: Aditya-Dahiya/visage
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/visage/book_solutions/chapter9.html"
  description: "Visualizing Information and Spatial Analysis with ggplot2 Extensions"
  twitter: true
  linkedin: true
  email: true
  mastodon: true
  facebook: true
---

In this chapter, I use **{ggplot2}** [@ggplot2-2] to produce equivalent maps, as produced by **{tmap}** [@tmap] in the textbook. In addition, I use **{cols4all}** [@cols4all] palettes for colour and fill scales.

```{r}
#| label: setup
#| eval: true

library(sf)        # Simple Features in R
library(terra)     # Handling rasters in R
library(tidyterra) # For plotting rasters in ggplot2
library(magrittr)  # Using pipes with raster objects
library(tidyverse) # All things tidy; Data Wrangling
library(spData)    # Spatial Datasets
library(patchwork) # Composing plots
library(gt)        # Display GT tables with R
```

### 9.1 Introduction

-   Cartography is a crucial aspect of geographic research, blending communication, detail, and creativity.
-   Static maps in R can be created using the [`plot()`](https://r.geocompx.org/adv-map.html#base-r-maps) function, but advanced cartography benefits from dedicated packages.
-   The chapter focuses in-depth on the [`tmap`](https://r-tmap.github.io/) package rather than multiple tools superficially.

```{r}
# install.packages("colorblindcheck")
# install.packages("cols4all")

# A nice way to pick colour palettes for maps etc.
# cols4all::c4a_gui()
cols4all::c4a_palettes() |> 
  as_tibble()

cols4all::c4a_series() |> 
  as_tibble()

cols4all::c4a_data()

cols4all::c4a_gui()

cols4all::scale_fill_binned_c4a_div()

cols4all::c4a(palette = "powerbi.highrise") |> 
  seecolor::print_color()
```

### 9.2 Static maps

-   Most common type of geo-computation output, stored as `.png` (raster) and `.pdf` (vector).
-   Base R's `plot()` is the fastest way to create static maps from [`sf`](https://r-spatial.github.io/sf/) or [`terra`](https://rspatial.github.io/terra/), ideal for quick visual checks.
-   [`tmap`](https://r-tmap.github.io/tmap/) package offers:
    -   Simple, ggplot2-like syntax.
    -   Static and interactive maps with `tmap_mode()`.
    -   Support for multiple spatial classes (`sf`, `terra`).

### 9.2.1 `tmap` basics

-   Grammar of graphics: Like [ggplot2](https://ggplot2.tidyverse.org/), [tmap](https://r-tmap.github.io/) follows a structured approach, separating input data from aesthetics (visual properties).
-   Basic structure: Uses `tm_shape()` to define the input dataset (vector or raster), followed by layer elements like `tm_fill()` and `tm_borders()`.
-   Layering approach:
    -   `tm_fill()`: Fills (multi)polygon areas.
    -   `tm_borders()`: Adds border outlines to (multi)polygons.
    -   `tm_polygons()`: Combines fill and border.
    -   `tm_lines()`: Draws lines for (multi)linestrings.
    -   `tm_symbols()`: Adds symbols for points, lines, and polygons.
    -   `tm_raster()`: Displays raster data.
    -   `tm_rgb()`: Handles multi-layer rasters.
    -   `tm_text()`: Adds text labels.
-   Layering operator: The `+` operator is used to add multiple layers.
-   Quick maps: `qtm()` provides a fast way to generate thematic maps (`qtm(nz)` ≈ `tm_shape(nz) + tm_fill() + tm_borders()`).
    -   Limitations of qtm(): Less control over aesthetics, so not covered in detail in this chapter.

### 9.2.2 Map objects

-   {tmap} allows storing maps as objects, enabling modifications and layer additions.
-   Use [`tm_polygons()`](https://r-tmap.github.io/tmap/reference/tm_polygons.html) to create a map object, combining `tm_fill()` and `tm_borders()`.
-   Stored maps can be plotted later by simply calling the object.
-   Additional layers are added using `+ tm_shape(new_obj)`, where `new_obj` represents a new spatial object.
-   Aesthetic functions apply to the most recently added shape until another is introduced.
-   Spatial objects can be manipulated with [`sf`](https://r-spatial.github.io/sf/), e.g., `st_union()`, `st_buffer()`, and `st_cast()`.
-   Multiple layers can be added, such as:
    -   Raster elevation (`tm_raster()`)
    -   Territorial waters (`tm_lines()`)
    -   High points (`tm_symbols()`)
-   [`tmap_arrange()`](https://r-tmap.github.io/tmap/reference/tmap_arrange.html) combines multiple `tmap` objects into a single visualization.
-   The `+` operator adds layers, but aesthetics are controlled within layer functions.

### 9.2.3 Visual variables

-   Default aesthetics in tmap:
    -   `tm_fill()` and `tm_symbols()` use gray shades.
    -   `tm_lines()` uses a continuous black line.
    -   Defaults can be overridden for customization.
-   Types of map aesthetics:
    -   Variable-dependent aesthetics (change with data).
    -   Fixed aesthetics (constant values).
-   Key aesthetic arguments in tmap:
    -   `fill`: Polygon fill color.
    -   `col`: Border, line, point, or raster color.
    -   `lwd`: Line width.
    -   `lty`: Line type.
    -   `size`: Symbol size.
    -   `shape`: Symbol shape.
    -   `fill_alpha`, `col_alpha`: Transparency for fill and border.
-   Applying aesthetics:
    -   Use a column name to map a variable. Pass a character string referring to a column name.
    -   Use a fixed value for constant aesthetics.
-   Additional arguments for visual variables:
    -   `.scale`: Controls representation on the map and legend.
    -   `.legend`: Customizes legend settings.
    -   `.free`: Defines whether each facet uses the same or different scales.

### 9.2.4 Scales

-   Scales define how values are visually represented in maps and legends, depending on the selected visual variable (e.g., `fill.scale`, `col.scale`, `size.scale`).
-   Default scale is [`tm_scale()`](https://r-tmap.github.io/tmap/reference/tm_scale.html), which auto-selects settings based on input data type (factor, numeric, integer).
-   Colour settings impact spatial variability; customization options include:
    -   `breaks`: manually set classification thresholds.
    -   `n`: define the number of bins.
    -   `values`: assign colour schemes (e.g., `"BuGn"`).
-   **Family of scale functions in *tmap*:**
    -   [`tm_scale_intervals()`](https://r-tmap.github.io/tmap/reference/tm_scale_intervals.html): splits values into intervals using predefined styles (`"pretty"`, `"equal"`, `"quantile"`, `"jenks"`, `"log10_pretty"`).
    -   [`tm_scale_continuous()`](https://r-tmap.github.io/tmap/reference/tm_scale_continuous.html): creates continuous colour fields, suitable for rasters.
    -   [`tm_scale_categorical()`](https://r-tmap.github.io/tmap/reference/tm_scale_categorical.html): assigns unique colours to categorical values.
-   Colour palettes are key for readability and should be carefully chosen:
    -   [`cols4all::c4a_gui()`](https://mtennekes.github.io/cols4all/) helps find suitable palettes.
    -   Default values for visual variables can be checked with [`tmap_options()`](https://r-tmap.github.io/tmap/reference/tmap_options.html).
-   Three main colour palette types:
    -   Categorical: distinct colours for unordered categories (e.g., land cover classes).
    -   Sequential: gradient from light to dark, for continuous numeric variables.
    -   Diverging: two sequential palettes meeting at a neutral reference point (e.g., temperature anomalies).
-   Key considerations for colour choices:
    -   Perceptibility: colours should match common associations (e.g., blue for water, green for vegetation).
    -   Accessibility: use colour-blind-friendly palettes where possible.
